---
title: "Final Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(broom)
library(stringr)
library(knitr)
library(nnet)
library(ggplot2)
```

### Introduction

### The Data

We will be using data that has extensive information on secondary school students in their math class. 

```{r}
data <- read.csv("data/student-mat.csv")


```
### Creation of New Variables

In order to provide more insight, we saw room to create informative variables based upon the given data. 

The given variables Medu and Fedu give information about the student's parents education history. Using this, we created a new variable "first_gen_college" that indicates if the student would be a first generation college student if they decided to pursue higher education. This will give more tangible and clear insight to how parental education impacts student's performance. 

```{r}
data <- data %>%
  mutate(first_gen_college = case_when(
    Medu < 4 & Fedu < 4 ~ "yes",
    TRUE ~"no"
  ))
```

Additionally, many variables are self reported ratings from the students on a scale of 1-5. We decided that instead factoring these variables so that scores of 1-3 would be "low" and scores of 4-5 would be "high" would be beneficial to our analysis as it would be more interpretable in context. 

```{r}
data <- data %>%
  mutate(famrel = case_when(
    famrel == 1 ~ "low",
    famrel == 2 ~ "low",
    famrel == 3 ~ "low",
    famrel == 4 ~ "high",
    famrel == 5 ~"high"
  ))

data <- data %>%
  mutate(freetime = case_when(
    freetime == 1 ~ "low",
    freetime == 2 ~ "low",
    freetime == 3 ~ "low",
    freetime == 4 ~ "high",
    freetime == 5 ~"high"
  ))

data <- data %>%
  mutate(goout = case_when(
    goout == 1 ~ "low",
    goout == 2 ~ "low",
    goout == 3 ~ "low",
    goout == 4 ~ "high",
    goout == 5 ~"high"
  ))

data <- data %>%
  mutate(Dalc = case_when(
    Dalc == 1 ~ "low",
    Dalc == 2 ~ "low",
    Dalc == 3 ~ "low",
    Dalc == 4 ~ "high",
    Dalc == 5 ~"high"
  ))

data <- data %>%
  mutate(Walc = case_when(
    Walc == 1 ~ "low",
    Walc == 2 ~ "low",
    Walc == 3 ~ "low",
    Walc == 4 ~ "high",
    Walc == 5 ~"high"
  ))

data <- data %>%
  mutate(health = case_when(
    health == 1 ~ "low",
    health == 2 ~ "low",
    health == 3 ~ "low",
    health == 4 ~ "high",
    health == 5 ~"high"
  ))
```

Additionally, using information from the famsup and internet variables, we created a variable called "stable_learning_env". If famsup is "yes" and internet is "yes", then stable_learning_env is "yes", otherwise "no".

```{r}
data <- data %>%
  mutate(stable_learning_env = case_when(
    internet =="yes" & famsup =="yes" ~"yes",
    TRUE ~"no"
  ))
```

Also, we created a new variable "high_freq_absent", which if absences >= 10 for a student, we considered them a highly frequent student.

```{r}
data <- data %>%
  mutate(high_freq_absent = case_when(
    absences >= 10 ~"yes",
    TRUE ~"no"
  ))
```

We also created a "failed" variable, which was "yes" if failures > 0, and "no" otherwise.

```{r}
data <- data %>%
  mutate(failed = case_when(
    failures > 0 ~"yes",
    TRUE ~"no"
  ))
```



### Exploratory Data Analysis 

```{r}
summary(data)
```

First, I will start off with univariate and bivariate plots of the response variable and key predictors I see being important. 

```{r}
data %>%
  filter(failed =="yes") %>%
  ggplot(aes(G3)) + 
  geom_histogram(stat = "count") +
  labs(title="Final Grade Distribution")

data %>%
  filter(failed =="yes") %>%
  ggplot(aes(G3)) + 
  geom_histogram(stat = "count") +
  labs(title="Final Grade Distribution")

data %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram()

data %>%
  keep(is.character) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram(stat="count")


```

Above we see that the response variable, G3, is pretty normally distributed, thus no transformation is necessary, 


```{r}
ggplot(data = data, aes(x = G3, y = first_gen_college, fill=first_gen_college)) +
  geom_boxplot() 
ggplot(data = data, aes(x = G3, y = Walc, fill = Walc)) +
  geom_boxplot() 
ggplot(data = data, aes(x = G3, y = famrel, fill = famrel)) +
  geom_boxplot() 
ggplot(data = data, aes(x = G3, y= sex, fill = sex)) +
  geom_boxplot() 
ggplot(data = data, aes(x = G3, y = high_freq_absent, fill = high_freq_absent)) +
  geom_boxplot() 
ggplot(data = data, aes(x = G3, y=failed, fill = failed)) +
  geom_boxplot() 
ggplot(data = data, aes(x = G3, y=romantic, fill = romantic)) +
  geom_boxplot() 
ggplot(data = data, aes(x = G3, y=internet, fill = internet)) +
  geom_boxplot() 
ggplot(data = data, aes(x = G3, y=goout, fill = goout)) +
  geom_boxplot() 

```

From the initial explorations above, we can see a few possible trends. Students who had at least one of the following traits: failed a class previously, were a highly frequent absent student, frequently went out, without internet, were frequent drinkers on the weekend, were in romantic relationships, and were first generation students, on average had lower final grades than their counterparts.


```{r}
names(data)
```

```{r}
num_cols <- unlist(lapply(data, is.numeric))
quant_vars <- data[,num_cols]
cor(quant_vars)

library(corr)
quant_vars %>% correlate() %>% network_plot(min_cor=0.2)
```


## Splitting data into training and testing sets

```{r}
data <- data %>%
  mutate(ord_g3 = factor(G3, ordered=T)
  )
data$ord_g3
attach(data)
set.seed(3)
train_ind <- sample(x = nrow(data), size = 0.8 * nrow(data))
test_ind_neg <- -train_ind
training <- data[train_ind, ]
testing <- data[test_ind_neg, ]
training
testing

```


## Multicategory ordinal logit model

```{r}

require(foreign)
require(nnet)
require(ggplot2)
require(reshape2)

mod1 <- polr(ord_g3 ~ failed + high_freq_absent + romantic + internet + goout + first_gen_college + Walc, data = training)
summary(mod1)
(ctable <- coef(summary(mod1)))
```

Calculate and store p-values:

```{r}
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = p))

```
Confidence intervals for parameter estimates:

```{r}
(ci <- confint(mod1))

confint.default(mod1) 
```

To get odds ratios:

```{r}
exp(coef(mod1))
exp(cbind(OR = coef(mod1), ci))

```

Testing set:

How do you evaluate it?? 

```{r}
pred.ord <- predict(mod1, testing)

```